---
title: "Analysis"
output: pdf_document
---
#TODO:
Drop zeros in DBP etc.
Do interaction terms
collinearity
Model selection
Show model to predict

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
# install.packages('SASxport')
library(SASxport)
library(dplyr)
library(tidyr)
health_demo <- read.xport('data/health_demo.XPT')

health_lab <- read.csv('data/health_lab.csv')
```
# Merge

```{r}
colnames(health_demo)
```

```{r}
colnames(health_lab)
```

```{r}
health_df <- merge(health_demo, health_lab, by.x="STUDY.ID", by.y="STUDY_ID",
                   all.x=FALSE, all.y=TRUE)
```

```{r}
health_df$RACE <- as.factor(health_df$RACE)
levels(health_df$RACE) <- c('Mexican American', 'Other Hispanic', 'Non-Hispanic White', 'Non-Hispanic Black', 'Non-Hispanic Asian', 'Other Race - Including Multi-Racial')

health_df$GENDER <- as.factor(health_df$GENDER)
levels(health_df$GENDER) <- c('male', 'female')
```

# Exercise: Fix education level

# Renaming variables

```{r}
health_df <- rename(health_df, Marital_Status = MARITALS) 
```

# "Dropping"/Keeping variables

```{r}

analysis_df <- health_df %>% select(STUDY.ID, GENDER, AGE, RACE, BMI, SBP, DBP, Triglycerides, LDL, Tot_chol, HDL, WBC, FastGlucose)

str(analysis_df)
  
```

```{r}
summary(analysis_df)
```
```{r}
nrow(analysis_df)
```


Note that we have ages 12-81.

```{r}
analysis_df <- analysis_df %>% filter((AGE >= 25) & (AGE <= 75))
nrow(analysis_df)
```

```{r}
summary(analysis_df)
```

 Note some extreme values that don't look right - data cleaning needed
 e.g. BMI = 61.70
 DBP = 0
 Triglycerides = 2141
 HDL = 226?
 WBC = 44.8 ?
 
 Note NAs

```{r}
hist(analysis_df$AGE)
```

```{r}
hist(analysis_df$AGE, breaks = 30)
```

```{r}
hist(analysis_df$BMI, breaks = 30) #slightly right skewed

hist(analysis_df$SBP, breaks = 30) # looks Normal
hist(analysis_df$DBP, breaks = 30) # looks Normal

hist((analysis_df$FastGlucose), breaks = 30) # go with log transformation instead. 
hist(log(analysis_df$FastGlucose), breaks = 30)

hist(analysis_df$Tot_chol, breaks = 30) #We see some values above 350ish in total chol.

hist(analysis_df$LDL, breaks = 30) # looks normal

hist(analysis_df$Triglycerides, breaks = 50) # NOT normal

hist(analysis_df$HDL, breaks = 30)

hist(analysis_df$WBC, breaks = 50) # some very high values
```

```{r drop_na}
analysis_df <- analysis_df %>% drop_na(SBP, DBP, AGE, BMI, HDL, FastGlucose, LDL)
```

```{r}
analysis_df$BMI_cat <- cut(analysis_df$BMI, breaks = c(0, 18.5, 25.0, 30.0, 35.0, 40, 100), labels = c("Underweight", "Normal", "Pre-obese", "Obesity I", "Obesity II", "Obesity III"), right = FALSE)

analysis_df$BMI_cat <- relevel(analysis_df$BMI_cat, "Normal") #this sets "normal" as the reference category.




#analysis_df$BMI_cat <- cut(analysis_df$BMI, breaks = c(0, 18.5, 25.0, 30.0, 100), labels = c("Underweight", "Normal", "Pre-obese", "Obese"), right = FALSE)

# analysis_df$BMI_cat <- relevel(analysis_df$BMI_cat, "Normal") #this sets "normal" as the reference category.

summary(analysis_df$BMI_cat)
```
```{r create_new_var}
analysis_df$HDL_LDL <- (analysis_df$HDL/analysis_df$LDL)
```


BLOOD PRESSURE CATEGORIES:
https://www.heart.org/-/media/data-import/downloadables/pe-abh-what-is-high-blood-pressure-ucm_300310.pdf

6. Let's create catgegories for variables we are intersted in.
```{r, eval=TRUE}
bp_cat <- function(sbp, dbp) {
 if (is.na(sbp) | is.na(dbp)){
   return(NA)
 }
  #if (sbp>=180 | dbp >=120) {
   # return("Hypertensive Crisis")
  #}
if (sbp>=140 | dbp >=90) {
    return("Hypertension Stage 2+")
  }

if ((sbp>=130 & sbp<=139) | (dbp >=80 & dbp <=89)) {
    return("Hypertension Stage 1")
  }

 if ((sbp>=120 & sbp<=129) &  dbp <80) {
    return("Elevated")
  }

  if (sbp < 120 & dbp <80) {
    return("Normal") 
  }
} 
```  


```{r, eval=TRUE}
# Converting Blood pressure into categories
analysis_df$bp_category <- mapply(bp_cat, analysis_df$SBP, analysis_df$DBP)

str(analysis_df$bp_category)
analysis_df$bp_category <- as.factor(analysis_df$bp_category)
analysis_df$bp_category <- relevel(analysis_df$bp_category, "Normal") #this sets "normal" as the reference category.

```
Do interactions




# Regression
Outcome: Total cholesterol
```{r}
full_model <- glm(FastGlucose ~ bp_category  + AGE  + BMI_cat + HDL_LDL , data = analysis_df)
summary(full_model)
```

Plotting residuals
```{r}
plot(resid(full_model))
```

```{r}
rst <- rstandard(full_model)
qqnorm(rst, ylab = "Standardized Residuals", 
       xlab = "Normal Scores")
qqline(rst)
```

```{r}
analysis_df$cooksd <- cooks.distance(full_model)
cooksd_cutoff <- 4/nrow(analysis_df)

plot(analysis_df$cooksd)
abline(cooksd_cutoff, 0)
```

Dropping outliers
```{r}
analysis_df2 <- analysis_df %>% filter(cooksd < cooksd_cutoff) %>% select(-cooksd)

full_model <- glm(FastGlucose ~ bp_category  + AGE  + BMI_cat + HDL_LDL , data = analysis_df2)
summary(full_model)
```

Plotting residuals
```{r}
plot(resid(full_model))
```

```{r}
rst <- rstandard(full_model)
qqnorm(rst, ylab = "Standardized Residuals", 
       xlab = "Normal Scores")
qqline(rst)
```
