---
title: "R Basics, Part 1: Data types and Functions"
teaching: 45
exercises: 20
questions:
- "How are different kinds of data represented in R?"
- "What is a variable in R?"
- "How do I test assertions about data?"
objectives:
- "Use and understand scalar data types"
- "Use and understand vectors and lists"
- "Define a variable"
- "Assign data to a variable"
- "Manage a workspace in an interactive R session"
- "Use mathematical and comparison operators"
- "Call functions"
- "Manage packages"
keypoints:
- "Use RStudio to write and run R programs."
- "R has the usual arithmetic operators and mathematical functions."
- "Use c() to construct a vector"
- "Use `<-` to assign values to variables."
- "Use `ls()` to list the variables in a program."
- "Use `rm()` to delete objects in a program."
- "Use `install.packages()` to install packages (libraries)."
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("03-")
```

# Variables in R

We can use *variables* to store data that we can then reference.  A variable has a name, which is like a label, and a value.

Let's look at an R expression that assigns a value to a variable:

```{r}
height_cm <- 140
```

This statement has three parts:

* An assignment operator `<-` which assigns the value on the right to the variable on the left.
* A variable name (`height_cm`)
* A value (`140`)

Let's take a look at what happened when you executed the statement `height_cm <- 140`.

In your Environment pane (usually in the upper right area of R Studio), you should now see a "Values" section.  This area is like a table of variable names and their values.  You should see `height_cm` listed with  a value of `140`.

If you now type the just name of the variable, `height_cm` at the console, R will **evaluate** the variable and print its current value.  Try it.

> ## Exercise 1
> 
> 1. Create a `height_inches` variable and set it to the value of `height_cm` divided by 2.54 (1 inch = 2.54 cm).
> 2. Then, change the value of `height_cm`.  Did `height_inches` recompute?  Why or why not?
>
> > ## Solution
> >
> > 1. `height_inches <- height_cm / 2.54`
> > 2. `height_inches` does not recompute.  It was set to the _*value*_ of `height_cm` divided by 2.54, which is a number.  It's not actually a formula or function that's tied to the `height_cm` variable.  We'll get to functions later!
> {: .solution}
{: .challenge}


## Variable names

Different programming languages have slightly different rules around what characters may be used in variable names.  In R, both numbers and letters can be used, as well as a dot (`.`) and an underscore (`_`).

> ## Exercise
> 
> 1. Can you determine whether a variable name may start with a number?  Can a variable name start with a period or an underscore?
>
> 2. Why do you think R allows an underscore (`_`) but not a hyphen (`-`) in variable names?
>
> 3. Can you determine whether variable names are case-sensitive?
>
> > ## Solution
> >
> > 1. something
> > 2. Something else
> > 3. And something else
> {: .solution}
{: .challenge}

## Variable Types

In our example above, the variables we used stored numeric values.  In real data, though, we have values that might be text, that might be true/false, might be empty, etc.

Let's try creating variables with values that are other types:

```{r}
study_id <- 'AB123X'
exposure <- TRUE
```


The idea here is that a variable not only stores data, but stores it as a particular type.

We can "ask" a variable what its type is, in R, using the `typeof()` function, like this:

```{r}
typeof(height_cm)
```

R tells us that `height_cm` is a "double".  What is a "double"?  "double" refers to a double-precision floating-point number.

If we check the type of `study_id`, we'll see that it's a `character` type.  `character` is the term in R that describes text data.  


If we check the type of `exposure`, we'll see that it's a `logical` type. `logical` is the term in R that describes data that can be either `TRUE` or `FALSE` (or `NA`).

Basic data types in R are:

* character
* numeric
* integer
* complex (for example, `4+3i`, where `i` represents the square root of -1)
* logical



# Functions

Earlier, we used a function, `typeof()`.  A function is a named set of code that can take zero or more **parameters** as input, performs some calculations or other processing, and (optionally) returns a value back.

Let's look at some examples of functions:

* Some mathematical functions you might use that can take a single value are functions like: `sqrt()`, `log()`, `exp()`:

```{r}
sqrt(415)
```
```{r}
x <- 524
log(x)
```
```{r}
exp(1)
```

Mathematical functions are only one of many types of functions we'll be using in R.  For example, we can use functions that work on strings:

```{r}
tolower('Capitalized Text')
```

There are, of course, functions that take no parameters, and functions that take more than one parameter:

```{r}
round(3.9472, digits = 2)
```

Notice in the above example that the `digits` parameter is optional.  If we leave it out, `round()` still works, because it has a **default** value for `digits` of 0:

```{r}
round(3.9472)
```

> ## Tip: Getting help on a function
>
> Nobody (not even the authors of this material) has memorized all of the functions in R.
> Looking up how to use a function is typically something you would do frequently while using R.
> 
> While you can certainly search the internet to learn more about an R function, RStudio provides several ways to read about a fucntion:
> 
> * At the RStudio Console, you can type `?` and then the name of a function.  For example, you might type `?round` and the Help pane, in the lower right area, will open up the documentation for that function.
>
> * If you're not sure which function you need, you can search for it by either:
>   * Typing in the search bar in the Help pane, or
>   * Typing `??` in the Console, followed by the text you want to search on.
>
{: .callout}

Later we'll create and use our own functions!


# Moving between data types: Coercion

Sometimes you have data as one type, but would like to "force" it to be a different type.  As an example, you might have "120" as text (character) data, but you'd like to treat it as a number, not text.  We'll see later that one place this can occur is when you're reading data into R from a file, and R makes some assumptions about the types of the data values.

R has a set of functions that start with `as.` which can help convert types:

```{r}
as.numeric("120.6")
```

```{r}
as.logical("True")
```

> ## Exercise
> 
> What happens when you try to coerce something that doesn't "fit" as the target type?  For example:
> 1. What happens when you try to coerce text with letters to a number?
> 2. What happens when you try to coerce text other than "true" or "false" to a logical value?
> 3. What happens when you try to coerce a floating-point number to an integer (and what function would you use to coerce to an integer?)
>
> > ## Solution
> > 1.  This results in an NA (note the warning as well)
> > ```{r}
> > as.numeric("abc")
> > ```
> > 2.  This also results in an NA
> > ```{r}
> > as.logical("maybe")
> > ```
> > 3.  You can use `as.integer()` and it will round the number to the nearest integer:
> > ```{r}
> > as.integer("11.82")
> > ```
> {: .solution}
{: .challenge}


# Logical expressions

Earlier we created a variable and set its value to `TRUE`.   We can also evaluate R expressions whose value will be `TRUE` or `FALSE` (or `NA`).

Although there are other types of expressions that evaluate to a logical value, logical expressions involving numerical quantities will often use operators like `<` (less than), `>` (greater than), `<=` (less than or equal to), `>=` (greater than or equal to), `==` (equal to), `!=` (not equal to)


> ## Discussion
> 
> 1. Why does R need a separate `==` operator to express equality?  Why not use `=`?
>
> > ## Solution
> >
> > In R, `=` is similar to `<-` in that, depending on the context, it can set the value on the left to the value on the right.  So an expression like `x == 3` would return `TRUE` or `FALSE` depending on whether or not `x` is equal to 3, but `x = 3` would _set_ `x` to 3.
> {: .solution}
{: .challenge}

We can also use logical operators in logical expressions, such as `&` (and), `|` (or), `!` (not).

As with mathematical expressions, using parentheses can help clarify your code and avoid having to remember the order of operations:

```{r}
heart_rate <- 72
blood_pressure_category <- "Normal"

(heart_rate < 96) & (blood_pressure_category == "Normal")
```



TODO:  Make the rest a new section


# Vectors and Atomic vectors

As R is built around working with statistics, an important construct in R is the **vector**.  A vector is a list or sequence of values.

note coercion that occurs - exercise re: which types win


# Functions, Revisited -- with Vectors




## TODO: Logicals - counting up values
TRUE + TRUE + FALSE + TRUE
yields 4
so... you can do things like
df$x <- c(4, 6, 9, 1, 0)
sum(df$x > 5)
