---
title: "Reading in Data"
teaching: 40
exercises: 5
questions:
- "How do I load in a data set?"
- "Once I've loaded in my data, how do I look at it?"
objectives:
- "Read a CSV file into a `data.frame`"
- "Read data files from other format: SPSS, SAS, Excel"
- "Explore a newly loaded data.frame -- number of rows, columns... preview before merging"
- "Understand the variable types of imported data"
- "Install R packages"
keypoints:
- "Learn to use read.* and related functions"
- "Learn to use str()"
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("01-")
```
# TODO For Fouzia: Describe what framingham dataset is. 

## Reading in Data

- This occurs in the analysis phase, after you've collected data (and/or using data from another source)
- Data should be in some sort of table format.  CSV, Excel, SAS, SPSS, Stata, etc.

R contains a number of built-in functions that can read in plain-text formats like CSV.  To read in formats associated with particular software packages, we'll need to use certain R packages (more on that in a bit).

Let's say that we have a CSV file called `health.csv`.  If we open the CSV file to look at it in a text reader (or in software like Excel), we see that it:
- has variable names in the first row (it doesn't have to - more on that soon #TODO)
- contains one observation per row

~~~
ID,First ,Last,Age,Gender,Blood Type,RBC Count,SBP,DPB
101,Dan,Chung,40,M,AB,7.35,121,80
102,Zara,Farooq,35,F,B,6.12,139,82
103,Sarah,Long,22,F,AB,6.89,125,81
104,Kurt,Woods,18,M,AB,5.71,126,82
105,Dan,Davis,45,M,,6.15,111,65
106,Matt,Van Jones,66,,A,4.7,,
107,John,Johnson,22,M,A,4.44,120,75
108,Veronica,Johnson,20,M,A,5.91,130,86
109,Cathy,Mallinson,32,F,,15.65,125,79
110,Mohammed,Ahmad,30,M,B,0.68,116,75
~~~

To read in the table, we can use the `read.csv()` function.  Let's see what what we get when we use it:

#TODO: Change the data, right now it's from a book
```{r}
read.csv('data/health.csv')
```

Where did it find the `data/health.csv` file?  #TODO

Let's notice a few things about what we got out:
- There seems to be a column to the left with numbering starting at 1
- other observations?

Next we're going to want to perform some analysis using this data, but to do that, we need to capture it in an R object somehow.

Let's try creating a variable called `health` and set it to whatever is returned by `read.csv()`:

```{r}
health <- read.csv('data/health.csv')
```

Running that line of code didn't print out the data table, like last time.  But in our Environment pane, we now see a new variable called `health`.

#TODO: other parameters in read.csv

> ## Exercise
> 
> 1. Click on `health` - see what happens.  Put questions here
>
> 2. Click on the `health` twisty.  What did you learn here?
>
> 3. Why do you think the blank values for Gender and Blood Type came through as blanks, but
>    the blanks in SBP and DBP came through as `NA`?
>
> 4. How would you find out what class of variable `health` is?  (answer:  class())
>
> > ## Solution
> >
> > ## 3.  Because of different variable types
> > ~~~
> > ## 4. `class(health)`
> > ~~~
> > {: .language-r}
> {: .solution}
{: .challenge}

Talk about `data.frame`

Let's look at a more realistic example of Public Health data.  One of the widely-used sources of public data in the U.S. is the National Health and Nutrition Examination Survey, NHANES.  Much of the data in NHANES is not provided in CSV format, and often contains many more variables and observations than in our first example.

In this case, we'll be working with **.XPT** files, which are in a format used by SAS.  If you work in public health, you'll most definitely be encountering data files in SAS "xport" format.  While XPT is a plain-text format, it's not terribly easy to read.

R's base packages don't include a function to read in XPT files, but we can install a "contributed" package called `SASxport`.

## A word about pacakges

Packages are sets of functions that # FIND A GOOD DEFINITION

*Installing* a package downloads the code to run the package onto your computer.  But we don't always want every package available in every project that we work on.

So, there's a second step involved to _load_ a package that you've installed.  More on that in a moment.

To install the `SASxport` package:

```{r}
#install.packages('SASxport')
```
Go to the **Packages** pane in RStudio.  Do you see that `SASxport` is now in the list of available packages?  You'll also notice that some of the packages are checked, and some, like SASxport, are not checked (yet).

TODO: Click on package name to get help (and what's the other way to get help for a package?)

You'll also notice Version numbers for all of the packages #Reproducibility

The next thing we need to do is load in the library for this particular project, using the `library()` function:

```{r, message=FALSE}
library(SASxport)
library(dplyr)
```

Notice that `SASxport` is now checked in the Packages pane.

We can open up the documentation page for `SASxport` in our Help pane by running `help(SASxport)`.  We learn there that the function we want to use is called `read.xport()`.  let's try it and see what we get.

`DEMO_I.XPT` contains demographic data for study subjects.

```{r}
framingham <- read.csv('data/frmgham2.csv')
```

After reading in our dataset into the `framingham` dataframe, we can inspect the shape and size of the data, such as: colname(df), head(df), tail(df)

```{r}
colnames(framingham)
```

# Variables of interest:
TCHOL
```{r}
framingham <- framingham %>% filter(PERIOD == 1) # should we do period 1 2 or 3? 
summary(framingham$AGE)
summary(framingham$SEX)
framingham$SEX <- as.factor(framingham$SEX)
# BMI
framingham$BMI_cat <- cut(framingham$BMI, breaks = c(0, 18.5, 25.0, 30.0, 35.0, 40, 100), labels = c("Underweight", "Normal", "Pre-obese", "Obesity I", "Obesity II", "Obesity III"), right = FALSE)
levels(framingham$BMI_cat)
summary(framingham$BMI_cat)# COMBINE OBESE II AND OBESE III CATEGORIES
#eventhough this categorical variable is already created by framingham, this will be a good exercise for class to do to create one themselves.  Perhaps then subset the dataframe with the one we create and the one that's already available to make sure it's created properly???
framingham$BMI_cat <- relevel(framingham$BMI_cat, "Normal") # sets "normal" as the reference category.
#NOTE
summary(framingham$BMI_cate) # this is already created by Framingham
```

# BP categories:
```{r}
#BLOOD PRESSURE:
framingham$SYSBP # Systolic blood pressure
framingham$DIABP # diastolic blood pressure

# Let's create catgegories for variables we are intersted in.
bp_category <- function(sbp, dbp) {
 if (is.na(sbp) | is.na(dbp)){
   return(NA)
 }
  #if (sbp>=180 | dbp >=120) {
   # return("Hypertensive Crisis")
  #}
if (sbp>=140 | dbp >=90) {
    return("Hypertension Stage 2+")
  }

if ((sbp>=130 & sbp<=139) | (dbp >=80 & dbp <=89)) {
    return("Hypertension Stage 1")
  }

 if ((sbp>=120 & sbp<=129) &  dbp <80) {
    return("Elevated")
  }

  if (sbp < 120 & dbp <80) {
    return("Normal") 
  }
} 

# Converting Blood pressure into categories
framingham$bp_cat <- mapply(bp_category, framingham$SYSBP, framingham$DIABP)
levels(factor(framingham$bp_cat))  # NOT Working
class(framingham$bp_cat)# shows it's a LIST so has to be converted to a dataframe:
framingham$bp_cat <- as.data.frame(lapply(framingham$bp_cat, unlist)) #this is not working

# Converting to factor 
framingham$bp_cat <- as.factor(framingham$bp_cat) # This conversion is not working b/c class of this variable is a LIST
framingham$bp_cat <- relevel(framingham$bp_cat, "Normal") #sets "normal" as reference.
str(framingham$bp_cat)  # Observe that it's a factor

table(framingham$bp_cat, useNA = "ifany")  # Inspect the values

barplot(table(framingham$bp_category)) #TODO reorder so it looks ordinal
``` 

# REGRESSIONS:
glm_tchol <- glm(TOTCHOL ~ BMI, data = framingham, family = gaussian(link = "identity")) # MAKE Categories of BMI
glm_tchol <- glm(TOTCHOL~ DIABETES, data = framingham, family = gaussian(link = "identity"))
glm_tchol <- glm(TOTCHOL ~ AGE, data = framingham, family = gaussian(link = "identity"))
glm_tchol <- glm(TOTCHOL ~ SEX, data = framingham, family = gaussian(link = "identity")) # has to be converted to a factor var
glm_tchol <- glm(TOTCHOL ~ CURSMOKE, data = framingham, family = gaussian(link = "identity")) # has to be factor
glm_tchol <- glm(TOTCHOL ~ CIGPDAY, data = framingham, family = gaussian(link = "identity"))
glm_tchol <- glm(TOTCHOL ~ SYSBP, data = framingham, family = gaussian(link = "identity"))
glm_tchol <- glm(TOTCHOL ~ DIABP, data = framingham, family = gaussian(link = "identity"))

glm_tchol <- glm(TOTCHOL ~ BMI_cat + AGE + SEX + DIABETES, data = framingham, family = gaussian(link = "identity"))
summary(glm_tchol)
exp(coef(glm_tchol))
```

### A word about reproducibility

Package versions - maybe put this as an open discussion question in the Exercises?  With some links to ways to solve reproducibility issues?

Platform-independent?  Maybe put this in the intro (R vs. software that only works on Windows, for example)

## Looking at variable types after reading in - Factors etc.

And how to use the `factorsAsStrings` parameter