---
title: "Selecting and Renaming Variables, and Subsetting"
teaching: 15
exercises: 2
questions:
- "How can I narrow down to just the variables of interest?"
- "How can I rename variables?"
objectives:
- "Learn how to scope a data frame down to just variables of interest"
- "Learn how to select variables by name using dplyr"
- "Learn how to rename variables"
keypoints:
- "Use `select()` to create a subset of a data frame based on columns"
- "Use `rename()` to rename columns"
- "Use `filter()` to filter rows"
source: Rmd
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("06-")
```

```{r, include=FALSE}
swan_df <- read.csv('data_out/swan_merge.csv')
```

## Creating a subset with fewer variables

Right now our data frame has hundreds of variables in columns, but for our analysis, we're only interested in analyzing data from a limited number of those variables.

Furthermore, we didn't pick the names for the variables.  They might be names like `TOTIDE26` or `GLUSSPE6` or `GLUSLOO6`.  Variable names like these are not at all uncommon when working with data from publicly available data sets, such as [NHANES](https://www.cdc.gov/nchs/nhanes/) or other [CDC](https://data.cdc.gov/) data sources, the [Framingham Heart Study](https://www.framinghamheartstudy.org/) or in the data we're using from the [SWAN](https://www.swanstudy.org/) study.   With variable names like these, it's especially important to download the codebook that explains each variable. 

And/or, the columns in the data sets might have names that can be difficult to work with in R.  For example, a column might be named `Study ID` or `Diastolic BP`, and names with spaces or other special characters might pose unique problems after the data is loaded into R.

So there are at least two steps we'd like to take at this point:
1. Narrow down to just certain variables
2. Rename variables

## Selecting variables

One of the most convenient ways to "wrangle" data frames is to use functionality that comes with the `dplyr` #TODO: Say more here about piping, etc.

dplyr's `select()` function narrows down to the variable names you specify as a comma-separated list of parameters.


```{r, message=FALSE}

library(dplyr)

analysis_swan_df <- swan_df %>% select(SWANID, AGE6, RACE, BMI6, 
                                       GLUCRES6, SMOKERE6, LDLRESU6, HDLRESU6, 
                                       CRPRESU6, DIABP16, SYSBP16, EXERCIS6)

str(analysis_swan_df)
  
```

`select()` also allows you to use other ways to express which columns to include or exclude.  For example, functions like `starts_with()`, `ends_with()`, `contains()` can be used with `select()`.

You might want to create a data frame by selecting all of the variables ending in "6", and you might do that with `select(ends_with("6"))`

`select()` also allows you to select all variables *except* a certain variable, using the hyphen (`-`).  For example, you could `select(-NAME, -BMI)` to select all variables except `NAME` and `BMI`.

>  ## Exercise
> 
> How might you create a new data frame from  `swan_df` containing just the id of each participant plus the glucose-related variables, which all start with "GLU"?
> 
> > ## Solution
> >
> > ```{r, message=FALSE}
> > glu_df <- swan_df %>% select(SWANID, starts_with("GLU"))
> > ```
> > {: .language-r}
> {: .solution}
{: .challenge}


## Renaming variables

Now that we've scoped our data frame down to just the variables we want to work with, let's rename a few, using the `rename()` function (also from `dplyr`).

Notice that each item in the parameter list is an expression of the form:

`new variable name = old variable name`


```{r}
analysis_swan_df <- analysis_swan_df %>% rename(Glucose = GLUCRES6, LDL = LDLRESU6, 
                           HDL = HDLRESU6, CRP =  CRPRESU6, DBP = DIABP16, SBP = SYSBP16,
                           Smoker = SMOKERE6, Exercise = EXERCIS6, Age = AGE6, BMI = BMI6)

str(analysis_swan_df)
```

> ## Exercise
> 
> Rename the `GLUSOST6` variable to `Glucosamine` in the `glu_df` data frame you created in the previous exercise.
> 
> > ## Solution
> >
> > ```{r, message=FALSE}
> > glu_df <- glu_df %>% rename(Glucosamine = GLUSOST6)
> > ```
> > {: .language-r}
> {: .solution}
{: .challenge}

```{r, include=FALSE}
write.csv(analysis_swan_df, 'data_out/analysis_swan_df_06.csv', row.names = FALSE)
```

#TODO:  As its own episode, show use of `filter()`
