---
title: "Merging and Joining Data"
keypoints:
- Use `merge()` to stack data
- Use `merge()` to join data
- Understand joins
objectives:
- Learn how to stack data frames (ex. different years' data)
- Learn how to join - inner, outer, etc.
- Explore after a join to verify results as expected
questions: How can I combine data from different sources?
source: Rmd
teaching: 45
exercises: 10
---

```{r, include=FALSE}
source("../bin/chunk-options.R")
knitr_fig_path("01-")
```

## Introduction goes here

Text

```{r, eval=TRUE}
library(SASxport)
library(tidyverse)
```

```{r, eval=TRUE}
demographics <- read.xport('data/DEMO_I.XPT')
bmi_df <- read.xport('data/BMX_I.XPT') 
bp_df <- read.xport('data/BPX_I.XPT')
cbc_df <- read.xport('data/CBC_I.XPT')
glu_df <- read.xport('data/GLU_I.XPT')
tg_df <- read.xport('data/TRIGLY_I.XPT')
hdl_df <- read.xport('data/HDL_I.XPT')
tc_df <- read.xport('data/TCHOL_I.XPT')
# smk_df <- read.xport('data/SMQ_I.XPT')
```

Now that we have all the files read in, let's merge.
From bmi_df we are keeping: BMXBMI
from bp_df: PBXPLS (pules), BPXSY1 (SBP), PBXDI1 (DPB),
tg_df : LBXTR (triglycerides mg/dL), LBDLDL (LDL mg/dL)
tc_df: LBXTC (total chol mg/dL)
hdl_df: LBDHDD (HDL mg/dL)
cbc_df: LBXWBCSI (WBCs 1000cells/uL)
glu_df: LBXGLU (glucose fasting mg/dL)

HERE WE NEED TO CONSIDER DPLYR FOR JOINING. #TODO

```{r, eval=TRUE}
merge_df <- merge(demographics, bmi_df[ ,c("SEQN", "BMXBMI")])
merge_df <- merge (merge_df, bp_df[ ,c("SEQN", "BPXSY1", "BPXDI1")])
merge_df <- merge(merge_df, tg_df[ , c("SEQN", "LBXTR", "LBDLDL")])
merge_df <- merge(merge_df, tc_df[ , c("SEQN", "LBXTC")])
merge_df <- merge(merge_df, hdl_df[ , c("SEQN", "LBDHDD")])
merge_df <- merge(merge_df, cbc_df[ , c("SEQN", "LBXWBCSI")])
merge_df <- merge(merge_df, glu_df[ , c("SEQN", "LBXGLU")])
```

```{r, eval=TRUE}
write.csv(merge_df, file = 'data_out/merge.csv', row.names = FALSE)
```

```{r, eval=TRUE}
merge_df <- read.csv("data_out/merge.csv")
```

Variables of interest:
BMI, BP, LDL, TC, HDL, GLU, AGE,
-We are interested in the association between TC and BMI categories. (Total chol (TC) = outcome and glucose (main predictor))

1. After merging, first look at NAs and summaries of variables.
```{r, eval=TRUE}
# Look at descriptive statistics of all independent variables

summary(merge_df$LBXTC) # total chol --> outcome variable.
sd(merge_df$LBXTC, na.rm = TRUE) # gives standard deviation; removes na values (na.rm=TRUE)
summary(merge_df$LBXGLU) # glucose
summary(merge_df$BMXBMI) # BMI
summary(merge_df$RIDAGEYR) #TODO exclude people <18yo now. 
```

2. Next, let's look at these variables in detail with graphs:
```{r, eval=TRUE}
hist(merge_df$LBXTC, breaks = 30, xlab = "Total cholesterol (TC) (mg/dl)",  main = "Distribution of Total Cholesterol") # looks normal overall
hist(merge_df$LBXGLU, breaks = 100) # right skewed
boxplot(merge_df$LBXGLU) # looking at outliers of IV
hist(merge_df$BMXBMI, breaks = 50) # slightly right skewed
boxplot(merge_df$BMXBMI) # looking at outliers of IV
hist(merge_df$RIDAGEYR, breaks = 50) # if you ignore >=80 category and <=10, it's normal looking. 

#TODO Create histograms etc of Blood pressure variables here (for continuous)
```

3. Let's drop anyone under the age of 25 b/c changes in TC may not be apparent in the younger age groups. 
```{r, eval=TRUE}
merge25_df <- merge_df %>% filter((RIDAGEYR>=25 & RIDAGEYR<80)) #& LBXGLU<=200)
boxplot(merge25_df$LBXGLU) # looking at outliers again in this new dataset
boxplot(merge25_df$BMXBMI) # looking at outliers again
```

4.Let's do scatter plots to see linear relationship between predictors and outcome variable:
```{r, eval=TRUE}
scatter.smooth(merge25_df$LBXGLU, merge25_df$LBXTC, xlab = "Glucose levels (mg/dl)", ylab = "Total Cholesterol (mg/dl)")
scatter.smooth(merge25_df$BMXBMI, merge25_df$LBXTC, xlab = "BMI", ylab = "Total Cholesterol (mg/dl)")
scatter.smooth(merge25_df$RIDAGEYR, merge25_df$LBXTC, xlab = "Age (years)", ylab = "Total Cholesterol (mg/dl)")
scatter.smooth(merge25_df$LBDHDD, merge25_df$LBXTC, xlab = "HDL", ylab = "Total Cholesterol (mg/dl)")
scatter.smooth(merge25_df$LBDLDL, merge25_df$LBXTC, xlab = "LDL", ylab = "Total Cholesterol (mg/dl)")
```

5. Let's do a correlation matrix now
```{r, eval=TRUE}
cor(merge25_df[ , c("LBXGLU", "BMXBMI", "RIDAGEYR", "LBDHDD", "LBDLDL")], use = "complete.obs") # everything looks like below 27% multicollinearity.
```

BLOOD PRESSURE CATEGORIES:
https://www.heart.org/-/media/data-import/downloadables/pe-abh-what-is-high-blood-pressure-ucm_300310.pdf

6. Let's create catgegories for variables we are intersted in.
```{r, eval=TRUE}
bp_cat <- function(sbp, dbp) {
 if (is.na(sbp) | is.na(dbp)){
   return(NA)
 }
  #if (sbp>=180 | dbp >=120) {
   # return("Hypertensive Crisis")
  #}
if (sbp>=140 | dbp >=90) {
    return("Hypertension Stage 2+")
  }

if ((sbp>=130 & sbp<=139) | (dbp >=80 & dbp <=89)) {
    return("Hypertension Stage 1")
  }

 if ((sbp>=120 & sbp<=129) &  dbp <80) {
    return("Elevated")
  }

  if (sbp < 120 & dbp <80) {
    return("Normal") 
  }
} 
```  


```{r, eval=TRUE}
# Converting Blood pressure into categories
# merge25_df <- merge25_df %>% mutate(bp_category = bp_cat(BPXSY1, BPXDI1))
merge25_df$bp_category <- mapply(bp_cat, merge25_df$BPXSY1, merge25_df$BPXDI1)
```

Let's first make sure that everything looks okay.  There are over 50 variables, so scrolling through the data frame in the viewer is not a great option.

```{r, eval=TRUE}
# See what the range of text vales is
unique(merge25_df$bp_category)

# Look at just the pertinent columns
just_bp <- merge25_df %>% select(BPXSY1, BPXDI1, bp_category) # check to make sure no NULL values and categories are correctly created

```

```{r, eval=TRUE}
# Converting to factor 
merge25_df$bp_category <- as.factor(merge25_df$bp_category) 
merge25_df$bp_category <- relevel(merge25_df$bp_category, "Normal") #sets "normal" as reference.
str(merge25_df$bp_category)  # Observe that it's a factor
table(merge25_df$bp_category, useNA = "ifany")  # Inspect the values

barplot(table(merge25_df$bp_category)) #TODO reorder so it looks ordinal
```

```{r, eval=TRUE}
# Convert BMI into categories
merge25_df$BMI_category <- cut(merge25_df$BMXBMI, breaks = c(0, 18.5, 25.0, 30.0, 35.0, 40, 100), labels = c("Underweight", "Normal", "Pre-obese", "Obesity I", "Obesity II", "Obesity III"), right = FALSE)

merge25_df$BMI_category <- relevel(merge25_df$BMI_category, "Normal") #this sets "normal" as the reference category.

summary(merge25_df$BMI_category)

```

NOTE: lm ignoes NAs (check this again though).  Make a note of this in the write up of the lesson. 

7. Simple linear regression models:
```{r, eval=TRUE}
#TODO: Set appropriate reference levels in factor variables (for BP and BMI)

# glucose
lm_gluc <- lm(LBXTC ~ LBXGLU, data = merge25_df) 
summary(lm_gluc)

# bmi
lm_bmi <- lm(LBXTC ~ BMXBMI, data = merge25_df)
summary(lm_bmi)

# bmi categorical
lm_bmicat <- lm(LBXTC ~ BMI_category, data = merge25_df)
summary(lm_bmicat)

# age
lm_age <- lm(LBXTC ~ RIDAGEYR, data = merge25_df)
summary(lm_age)

# hdl
lm_hdl <- lm(LBXTC ~ LBDHDD, data = merge25_df)
summary(lm_hdl)

# ldl
lm_ldl <- lm(LBXTC ~ LBDLDL, data = merge25_df)
summary(lm_ldl)

# BP categorical
lm_bp <- lm(LBXTC ~ bp_category, data = merge25_df)
summary(lm_bp)
```

#TODO  INTERACTION TERMS, look at literature to see which ones need to be created. 

Next: Just creating one big model just to see for now.
```{r, eval=TRUE}
lm_model <- lm(LBXTC ~ LBXGLU + BMI_category + LBDHDD + LBDLDL + RIDAGEYR + bp_category, data = merge25_df)
summary(lm_model)
```


